---
title: "Effects of hydropower regulation on timing of emergence in Trichoptera"
author: "Christianne Dalsbotten Solvåg"
output: html_document
---

# SET-UP ------------

```{r Settings}

settings = list(
  update_R_now = FALSE,
  get_citations_now = FALSE,
  remove_microtrichoptera = TRUE
)


# Load colour palette and preferred figure format
source("figure_formats_and_colour_palette.R")

```

```{r Update R}
if (settings$update_R_now){
  
  # Source - https://stackoverflow.com/a/29050116
  # Posted by Borealis, modified by community. See post 'Timeline' for change history
  # Retrieved 2026-05-04, License - CC BY-SA 3.0
  
  # installing/loading the package:
  if(!require(installr)) {
  install.packages("installr"); require(installr)} #load / install+load installr
  
  # using the package:
  updateR() # this will start the updating process of your R installation.  It will check for newer versions, and if one is available, will guide you through the decisions you'd need to make.
}

```

```{r Load packages}

## Custom install & load function (From OS Course, Erik Kusch)
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, repos = "http://cran.us.r-project.org")
  }
  require(x, character.only = TRUE)
}
## names of packages we want installed (if not installed yet) and loaded
package_vec <- c(
  "rmarkdown", # r markdown
  "here", # package for locating and storing files with relative paths
  "dplyr", # data manipulation
  "knitr", # for rmarkdown table visualisations
  "ggplot2", # for visualistion
  "ggpubr", # publication ready plots
  "stringr", # for string operations
  "readxl", # for reading xlsx-files
  "svglite", # for saving plots as SVG-files
  "tidyr",
  "MASS", # modern applied statistics with S
  "lme4", # Mixed-effects models
  "MuMIn", # Model selection, AICc, dredge
  "ordinal", # CLMs and CLMMs
  "ggtext", # markdown in ggplots
  "scales"
)

## executing install & load for each package
sapply(package_vec, install.load.package)

```

```{r Cite packages}
if (settings$get_citations_now){
  version$version.string
  citation() # R base
  package_vec %>% sapply(citation) %>% sapply(toBibtex)
}
```



# LOAD DATA ----------------

```{r Load data on events + taxon list}

# Info workbook, events sheet: Main sheet where all the corresponding info about localities, sampling methods etc should be added.
events_df <- readxl::read_excel(path = here::here("data","source_data","raw","info_fieldstudy_phenology.xlsx"), sheet = 1) # Can also specify sheet name, i.e., sheet = "events"

# Change "round" to "sample" 
events_df$internal_eventID <- str_replace_all(events_df$internal_eventID, "round","sample")

# Taxon list with info
taxonlist_df <- readxl::read_excel(path = here::here("data","source_data","raw","taxonlist_fieldstudy_PhD.xlsx"), sheet = 1)
taxonlist_trich_df <- filter(taxonlist_df, order == "Trichoptera")

```

```{r Lake types (function)}

# Lake names
lake_names <- c("Barsetvatnet","Kilvatnet","Roksetvatnet","Gjoljavatnet","Stor-Drakstsjoen","Storvatnet")

# Lake types
lake_status <- function(lake_name){
  case_when(lake_name %in% c("Barsetvatnet","Kilvatnet","Roksetvatnet") ~ "Control",
            lake_name %in% c("Gjoljavatnet","Stor-Drakstsjoen","Storvatnet") ~ "Regulated",
            .default = c("Unknown group"))
}


```

```{r Load data from kick samples (larvae)}

# Load files ------------

# Import all occurrence files that are in the folder "occurrences"
kicksample_occurrences_df <- list.files(path = here::here("data","source_data","raw","kicksample_occurrences"), # create a list of all excel files in folder
                             pattern = "*xlsx",
                             full.names = TRUE) %>%
  lapply(read_excel, sheet = 1) %>% # read all the files in the list
  bind_rows() # Merge into single dataframe

# Remove empty rows (where id is NA), result of adding empty rows in excel
kicksample_occurrences_df <- kicksample_occurrences_df[!is.na(kicksample_occurrences_df$internal_parentEventID),] 

# Remove unnecessary columns
kicksample_occurrences_df <- kicksample_occurrences_df %>% 
  dplyr::select(internal_eventID,scientificName,individualCount) # remove "internal_parentEventID" and "order" columns (should not have been in excel)

# Change "round" to "sample" 
kicksample_occurrences_df$internal_eventID <- str_replace_all(kicksample_occurrences_df$internal_eventID, "round","sample")



#### Add info ------------

# Add locality/event info to occurrence file
kicksample_occurrences_df <- merge(x = kicksample_occurrences_df, y = events_df,
                        by = "internal_eventID", all.x = TRUE) 
  # ideally, here we will only have one key column, internal_eventID

# Add taxon info to occurrence file
kicksample_occurrences_df <- merge(x = kicksample_occurrences_df, y = taxonlist_df,
                        by = "scientificName", all.x = TRUE)




#### Tidy data --------------

# One replicate (chosen at random) was processed for each event, except the first one in 2023, where three replicates were processed.
# For lake Stor-Drakstsjøen, two additional replicates were semi-processed (EPT was counted, but not identified to species level) for the date 24.05.2023. 
# This was done early in the sorting process, when the time available for labwork was being assessed. Further, having multiple replicates for the same time point allowed us to compare how similar replicates were in terms of species composition and abundance. To avoid different sample sizes for the different lakes, the two extra replicates are removed from the dataset.

# Remove additional replicates
kicksample_occurrences_df <- kicksample_occurrences_df %>%
  dplyr::filter(!internal_eventID %in% c("Stor-Drakstsjoen_24_05_2023_sample_4",
                                       "Stor-Drakstsjoen_24_05_2023_sample_5"))

# A related MSc project collected kick samples also in for one sampling event in September 2023. Not all of these were processed, so we remove these to avoid imbalanced data set.

# Remove incomplete autumn data 
kicksample_occurrences_df <- kicksample_occurrences_df %>%
  filter(month < 8) 

# Ø -> O 
kicksample_occurrences_df <- kicksample_occurrences_df %>%
  mutate(internal_parentEventID = iconv(internal_parentEventID,"UTF-8", "ASCII//TRANSLIT", sub = "o"),
         internal_eventID = iconv(internal_eventID,"UTF-8", "ASCII//TRANSLIT", sub = "o"),
         locality = iconv(locality,"UTF-8", "ASCII//TRANSLIT", sub = "o"))

# Create regulation status column 
kicksample_occurrences_df <- kicksample_occurrences_df %>% 
  mutate(Status = lake_status(locality))

# Make locality a factor and set level order 
kicksample_occurrences_df$locality <- factor(kicksample_occurrences_df$locality, 
                                  levels =  c("Barsetvatnet","Kilvatnet","Roksetvatnet","Gjoljavatnet","Stor-Drakstsjoen","Storvatnet"))

# Make year a factor and set level order
kicksample_occurrences_df$year <- factor(kicksample_occurrences_df$year, 
                              levels = c(2023, 2024))



# As kick sample occurences contain taxa other than Trichoptera, we subset the dataframe.
kicksample_trich_occurrences_df <- kicksample_occurrences_df %>% filter(order == "Trichoptera")

# Most of the time, we are only interested in larval individuals that have been identified to species level.
kicksample_trich_occurrences_speciesonly_df <- filter(kicksample_trich_occurrences_df, taxonRank == "SPECIES")



```

```{r Load data from Malaise traps (adults)}

## Load raw Trichoptera data from Malaise traps
malaise_occurrences_df <- list.files(path = "data/source_data/raw/malaise_data_occurences/", # go to folder
                         pattern = "*.xlsx", # look for Excel-files
                         full.names = TRUE) %>% # include path in filename
  lapply(read_excel, sheet = "occurrences") %>% # read all the files in the list 
  bind_rows() %>% # Merge into single dataframe
  filter(complete.cases(.)) # remove incomplete cases, i.e. rows with NA

# Remove unnecessary/duplicate columns
malaise_occurrences_df <- malaise_occurrences_df %>% 
  dplyr::select(internal_eventID,scientificName,individualCount) # remove "internal_parentEventID" and "order" columns (should not have been in excel)

# Change "round" to "sample" 
malaise_occurrences_df$internal_eventID <- str_replace_all(malaise_occurrences_df$internal_eventID, "round","sample")


# Add locality/event info to occurrence file
malaise_occurrences_df <- merge(x = malaise_occurrences_df, y = events_df,
                    by = c("internal_eventID"), all.x = TRUE) 


# Update to correct protocol
malaise_occurrences_df$samplingProtocol <- "Malaise trap"
malaise_occurrences_df$samplingEffort <- "Single Malaise trap, operated for approximately two weeks"

# Add taxon info to occurrence file
malaise_occurrences_df$scientificName <- str_replace_all(malaise_occurrences_df$scientificName, "\\s", " ") # replaces a few instances with a white space that is not normal space
malaise_occurrences_df <- merge(x = malaise_occurrences_df, y = taxonlist_df,
                    by = "scientificName", all.x = TRUE)


# Change ø to o
malaise_occurrences_df <- malaise_occurrences_df %>%
  mutate(locality = iconv(locality,"UTF-8", "ASCII//TRANSLIT", sub = "o"),
         internal_eventID = iconv(internal_eventID,"UTF-8", "ASCII//TRANSLIT", sub = "o"))


# Create regulation status column
malaise_occurrences_df <- malaise_occurrences_df %>% 
  mutate(Status = lake_status(locality))


# Make locality a factor and set level order
malaise_occurrences_df$locality <- factor(malaise_occurrences_df$locality, 
                                  levels = lake_names)



# Make year a factor and set level order
malaise_occurrences_df$year <- factor(malaise_occurrences_df$year, 
                          levels = c(2023, 2024))


```

```{r Remove data on micro-caddisfly (if true)}

if (settings$remove_microtrichoptera){
  # Larva data
  kicksample_occurrences_df <- kicksample_occurrences_df %>% filter(genus != "Oxyethira", family != "Hydroptilidae")
  kicksample_trich_occurrences_df <- kicksample_trich_occurrences_df %>% filter(genus != "Oxyethira", family != "Hydroptilidae")
  kicksample_trich_occurrences_speciesonly_df <- kicksample_trich_occurrences_speciesonly_df %>% filter(genus != "Oxyethira", family != "Hydroptilidae")
  
  # Adult data
  malaise_occurrences_df <- malaise_occurrences_df %>% filter(genus != "Oxyethira", family != "Hydroptilidae")
}

```

# DATA OVERVIEW ----------------

## Species lists

```{r Species list}

all_observations_species <- bind_rows(kicksample_trich_occurrences_speciesonly_df, malaise_occurrences_df)

specieslist_long <- all_observations_species %>% group_by(scientificName, samplingProtocol, Status) %>% summarise(n = sum(individualCount))
unique(specieslist_long$scientificName)

specieslist_wide_ctr <- spread(filter(specieslist_long, Status == "Control"), samplingProtocol, n)
specieslist_wide_reg <- spread(filter(specieslist_long, Status == "Regulated"), samplingProtocol, n)

specieslist_by_status_x_method <- full_join(specieslist_wide_ctr, specieslist_wide_reg, by = "scientificName", suffix = c(".ctrl", ".reg")) %>%
  dplyr::select("scientificName", "Kick.reg" = "Rot (1 min).reg", "Malaise.reg" = "Malaise trap.reg", "Kick.ctrl" = "Rot (1 min).ctrl", "Malaise.ctrl" = "Malaise trap.ctrl") %>% arrange(scientificName)

rm(specieslist_long, specieslist_wide_ctr, specieslist_wide_reg)

```

```{r Species list - all taxon ranks}

all_observations <- bind_rows(kicksample_trich_occurrences_df, malaise_occurrences_df)

taxon_list_long <- all_observations %>% group_by(taxonRank, family, genus, scientificName, samplingProtocol, Status) %>% summarise(n = sum(individualCount))
unique(taxon_list_long$scientificName)

taxon_list_wide_ctr <- spread(filter(taxon_list_long, Status == "Control"), samplingProtocol, n)
taxon_list_wide_reg <- spread(filter(taxon_list_long, Status == "Regulated"), samplingProtocol, n)

taxonlist_by_status_x_method <- full_join(taxon_list_wide_ctr, taxon_list_wide_reg, by = c("taxonRank","family","scientificName"), suffix = c(".ctrl", ".reg")) %>% dplyr::select("taxonRank","scientificName", "Kick.reg" = "Rot (1 min).reg", "Malaise.reg" = "Malaise trap.reg", "Kick.ctrl" = "Rot (1 min).ctrl", "Malaise.ctrl" = "Malaise trap.ctrl") %>% arrange(family, scientificName)

rm(taxon_list_long, taxon_list_wide_ctr, taxon_list_wide_reg)

```


## Quick stats for adults (Malaise data)

```{r Quick stats for Malaise data}
print("---- Stats across all Malaise ----")
paste("Number of individuals (ID): ", sum(malaise_occurrences_df$individualCount)) %>% print()
paste("Number of species: ", length(unique(filter(malaise_occurrences_df, taxonRank == "SPECIES")$scientificName))) %>% print()

print("---- Stats for reg Malaise ----")
paste("Number of individuals (ID): ", sum(filter(malaise_occurrences_df, Status == "Regulated")$individualCount)) %>% print()
paste("Number of species: ", length(unique(filter(malaise_occurrences_df, taxonRank == "SPECIES", Status == "Regulated")$scientificName))) %>% print()


print("---- Stats for ctrl Malaise ----")
paste("Number of individuals (ID): ", sum(filter(malaise_occurrences_df, Status == "Control")$individualCount)) %>% print()
paste("Number of species: ", length(unique(filter(malaise_occurrences_df, taxonRank == "SPECIES", Status == "Control")$scientificName))) %>% print()


```



# ANALYSIS 1: ADULT TEMPORAL DISTRIBUTION ----------------

Y = period an adult individual is detected, explained by year and status (random effects: locality and species). Data is at individual level.

```{r Adult presence (only species in Malaise (no filtering, no absence))}

# One line per individual
adult_individuals_trichoptera_df <- malaise_occurrences_df[rep(row.names(malaise_occurrences_df), times = malaise_occurrences_df$individualCount), ]

# Keep only most relevant columns
adult_individuals_trichoptera_df <- adult_individuals_trichoptera_df %>% transmute(year, periodNumber, Status, locality, scientificName, genus, taxonRank)

# Categorical emergence time
adult_individuals_trichoptera_df$period <-  case_when(
  adult_individuals_trichoptera_df$periodNumber == 1 ~ "Period1",
  adult_individuals_trichoptera_df$periodNumber == 2 ~ "Period2",
  adult_individuals_trichoptera_df$periodNumber == 3 ~ "Period3",
  adult_individuals_trichoptera_df$periodNumber == 4 ~ "Period4") %>% factor(levels = c("Period1", "Period2", "Period3", "Period4"))
 

adult_individuals_trichoptera_df$locality <- factor(adult_individuals_trichoptera_df$locality)
adult_individuals_trichoptera_df$Status <- factor(adult_individuals_trichoptera_df$Status)
adult_individuals_trichoptera_df$genus <- factor(adult_individuals_trichoptera_df$genus)



```

```{r Count table for adult Trichoptera (identified individuals only)}

adultcount_table_long <- malaise_occurrences_df %>% 
  group_by(year, Status, locality, periodNumber) %>%
  summarise(sum_n = sum(individualCount)) %>%
  ungroup() %>%
  complete(year, locality, periodNumber = c(1,2,3,4)) %>%
  transmute(year, Status, locality, periodNumber, sum_n) %>% # sets order of columns
  mutate(Status = lake_status(locality))

# Pivot to wide format (each period number as a column)
adultcount_table_wide <- adultcount_table_long %>%
  pivot_wider(names_from = periodNumber, values_from = sum_n, names_prefix = "Period")

adultcount_table_wide

# Quick stats
paste("Number of individuals: ", nrow(adult_individuals_trichoptera_df)) %>% print()
paste("Number of species: ", length(unique(adult_individuals_trichoptera_df$scientificName))) %>% print()

```

```{r A1: CLMMs for adult emergence time}

# Do a cumulative logit model of adult presence


A1_mod_0_null <- ordinal::clmm(period ~ 1 + (1|scientificName) + (1|locality),
                                           data = adult_individuals_trichoptera_df, 
                                           link = "logit")

A1_mod_1_year <- ordinal::clmm(period ~ year + (1|scientificName) + (1|locality),
                                           data = adult_individuals_trichoptera_df, 
                                           link = "logit")
A1_mod_2_status <- ordinal::clmm(period ~ Status + (1|scientificName) + (1|locality),
                                           data = adult_individuals_trichoptera_df, 
                                           link = "logit")
A1_mod_3_year_status <- ordinal::clmm(period ~ year + Status + (1|scientificName) + (1|locality),
                                           data = adult_individuals_trichoptera_df, 
                                           link = "logit")
A1_mod_4_interaction <- ordinal::clmm(period ~ year * Status + (1|scientificName) + (1|locality),
                                           data = adult_individuals_trichoptera_df, 
                                           link = "logit") 
# A1_mod_4_interaction gives warning: Hessian is numerically singular: parameters are not uniquely determined \\ In addition: Absolute convergence criterion was met, but relative criterion was not met


A1_candidate_models <- list(
    A1_mod_0_null,
    A1_mod_1_year,
    A1_mod_2_status,
    A1_mod_3_year_status,
    A1_mod_4_interaction
  )
MuMIn::model.sel(A1_candidate_models)

A1_favourite_model <- MuMIn::get.models(MuMIn::model.sel(A1_candidate_models), subset = 1)[[1]]
A1_favourite_model

summary(A1_favourite_model)
confint(A1_favourite_model)

```

```{r A1: All model summaries}
lapply(A1_candidate_models, summary)

```

```{r A1: Calculating predictions}

# A1_favourite_model has two predictors: Status and Year

# Create a new data frame for prediction (e.g., all combinations of Status and year)
A1_predictions <- expand.grid(
  j_period = levels(adult_individuals_trichoptera_df$period),
  x1_Status = levels(adult_individuals_trichoptera_df$Status),
  x2_year = levels(adult_individuals_trichoptera_df$year)
)

# Get values for thresholds (alpha) and predictors (beta)
A1_predictions <- A1_predictions %>% mutate(
  alpha_j = case_when(
    j_period == "Period2" ~ A1_favourite_model$alpha[["Period2|Period3"]],
    j_period == "Period3" ~ A1_favourite_model$alpha[["Period3|Period4"]]
  ),
  beta_x1 = case_when(
    x1_Status == "Control" ~ 0,
    x1_Status == "Regulated" ~ 1*A1_favourite_model$beta[["StatusRegulated"]]
  ),
  beta_x2 = case_when(
    x2_year == 2023 ~ 0,
    x2_year == 2024 ~ 1*A1_favourite_model$beta[["year2024"]]
  )
)

# Calculate logit(P(Y<=j))
A1_predictions <- A1_predictions %>% mutate(logit_P_Y_leq_j = case_when(
  j_period == "Period1" ~ NA,
  .default = alpha_j - beta_x1 - beta_x2))

# Calculate P(Y<=j)
A1_predictions <- A1_predictions %>% mutate(exp_logit = exp(logit_P_Y_leq_j))
A1_predictions <- A1_predictions %>% mutate(P_Y_leq_j = case_when(
  j_period == "Period1" ~ 0.0,
  j_period == "Period4" ~ 1.0,  # Assuming Period 4 is the last
  .default = exp_logit/(1+exp_logit)))

# Calculate P(Y=j)
A1_predictions <- A1_predictions %>% mutate(P_Y_eq_j = case_when(
  j_period == "Period1" ~ 0,
  .default = P_Y_leq_j-lag(P_Y_leq_j)))

```

```{r A1: Prediction uncertainty via simulation}
# Simulate coefficient uncertainty for the fitted CLMM and derive prediction intervals.
# This captures uncertainty in thresholds and fixed effects for the marginal predictions.

# For ordinal CLMM, extract both thresholds (alpha) and fixed effects (beta)
coefs <- c(A1_favourite_model$alpha, A1_favourite_model$beta)

vc    <- vcov(A1_favourite_model)

# Ensure vcov is subsetted to match coefs
vc_matched <- vc[names(coefs), names(coefs)]

set.seed(2026)
n_sims <- 2000
sim_coef_mat <- MASS::mvrnorm(n = n_sims, mu = coefs, Sigma = vc_matched)

simulated_A1_predictions <- lapply(seq_len(n_sims), function(sim_i) {
  theta <- sim_coef_mat[sim_i, ]

  A1_predictions %>%
    mutate(
      alpha_j = case_when(
        j_period == "Period2" ~ theta[["Period2|Period3"]],
        j_period == "Period3" ~ theta[["Period3|Period4"]],
      ),
      beta_x1 = case_when(
        x1_Status == "Control" ~ 0,
        x1_Status == "Regulated" ~ theta[["StatusRegulated"]]
      ),
      beta_x2 = case_when(
        x2_year == 2023 ~ 0,
        x2_year == 2024 ~ theta[["year2024"]]
      ),
      logit_P_Y_leq_j = alpha_j - beta_x1 - beta_x2,
      P_Y_leq_j = case_when(
        j_period == "Period1" ~ 0,
        j_period == "Period2" ~ plogis(logit_P_Y_leq_j),
        j_period == "Period3" ~ plogis(logit_P_Y_leq_j),
        j_period == "Period4" ~ 1
      )
    ) %>%
    group_by(x1_Status, x2_year) %>%
    arrange(j_period, .by_group = TRUE) %>%
    mutate(P_Y_eq_j = P_Y_leq_j - lag(P_Y_leq_j, default = 0)) %>%
    ungroup() %>%
    mutate(sim = sim_i)
})

simulated_A1_predictions <- dplyr::bind_rows(simulated_A1_predictions)

A1_prediction_intervals <- simulated_A1_predictions %>%
  group_by(j_period, x1_Status, x2_year) %>%
  summarise(
    P_Y_eq_j_mean = mean(P_Y_eq_j),
    P_Y_eq_j_se   = sd(P_Y_eq_j),
    P_Y_eq_j_lo95   = quantile(P_Y_eq_j, 0.025),
    P_Y_eq_j_hi95   = quantile(P_Y_eq_j, 0.975),
    P_Y_eq_j_lo68   = quantile(P_Y_eq_j, 0.5-0.341),
    P_Y_eq_j_hi68   = quantile(P_Y_eq_j, 0.5+0.341),
    P_Y_leq_j_mean = mean(P_Y_leq_j),
    P_Y_leq_j_se   = sd(P_Y_leq_j),
    P_Y_leq_j_lo95   = quantile(P_Y_leq_j, 0.025),
    P_Y_leq_j_hi95   = quantile(P_Y_leq_j, 0.975),
    P_Y_leq_j_lo68   = quantile(P_Y_leq_j, 0.5-0.341),
    P_Y_leq_j_hi68   = quantile(P_Y_leq_j, 0.5+0.341),
    .groups = "drop"
  )

A1_predictions <- A1_predictions %>%
  left_join(A1_prediction_intervals,
            by = c("j_period", "x1_Status", "x2_year"))


```

```{r A1: Visualize predictions with uncertainty}
# Line-and-dot plot with ribbon and error bars for P(Y=j) and cumulative P(Y<=j).

p_eq_A1 <- ggplot(A1_predictions, aes(x = j_period, y = P_Y_eq_j_mean, group = x1_Status, color = x1_Status)) +
  geom_errorbar(aes(ymin = P_Y_eq_j_lo95, ymax = P_Y_eq_j_hi95), width = 0.2, position = position_dodge(width = 0.3)) +
  geom_line(position = position_dodge(width = 0.3), linewidth = 1) +
  geom_point(position = position_dodge(width = 0.3), size = 2) +
  facet_wrap(~x2_year, nrow = 2) +
  labs(
       x = "Sampling period (j)",
       y = "P(observation time = j)") +
  scale_color_manual(name = "Status", values = status_colours) +
  scale_fill_manual(name = "Status", values = status_colours) +
  scale_x_discrete(labels = c("Period1" = "1", "Period2" = "2", "Period3" = "3", "Period4" = "4", "Later" = "Later")) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank())

p_leq_A1 <- ggplot(A1_predictions, aes(x = j_period, y = P_Y_leq_j_mean, group = x1_Status, color = x1_Status)) +
  geom_errorbar(aes(ymin = P_Y_leq_j_lo95, ymax = P_Y_leq_j_hi95), width = 0.2, position = position_dodge(width = 0.3)) +
  geom_line(position = position_dodge(width = 0.3), linewidth = 1) +
  geom_point(position = position_dodge(width = 0.3), size = 2) +
  facet_wrap(~x2_year, nrow = 2) +
  labs(
       x = "Sampling period (j)",
       y = "P(observation time \u2264 j)") +
  scale_color_manual(name = "Status", values = status_colours) +
  scale_fill_manual(name = "Status", values = status_colours) +
  scale_x_discrete(labels = c("Period1" = "1", "Period2" = "2", "Period3" = "3", "Period4" = "4", "Later" = "Later")) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank())


p_eq_A1
p_leq_A1

for (figformat in preferred_formats){
  # NOTE! For some reason "less or equal to" symbol is displayed as = in pdf (svg and png OK)
  ggsave(paste("figures/results/adultobs_peq.", figformat, sep = ""), plot = p_eq_A1, height = 10, width = col_width_dbl, units = "cm")
  ggsave(paste("figures/results/adultobs_pleq.", figformat, sep = ""), , plot = p_leq_A1, height = 10, width = col_width_dbl, units = "cm")
}

```

# ANALYSIS 2: FIRST EMERGENCE ------------

Criteria/"rules": **For each lake, ...**

1\. If a species is caught in the Malaise trap a certain year, "first emergence" will be set to the **first period they are observed that year**

2\. If a species has **not** been caught in the Malaise trap in a certain year, but is observed (in kick-samples) **two or more times in the focal year**, "first emergence" will be set to **"Later"**

3\. If a species has **not** been caught in the Malaise trap in a certain year, but is observed (in kick-samples) **three or more times across years**, "first emergence" will be set to **"Later"**

(Note that "times" refer to periods, not individuals.)

```{r A2: Rules for first emergence}

kick_periods_year <- kicksample_trich_occurrences_speciesonly_df %>%
  distinct(year, locality, scientificName, periodNumber) %>%
  group_by(year, locality, scientificName) %>%
  summarize(kick_periods_year = n(), .groups = "drop")

kick_periods_across <- kicksample_trich_occurrences_speciesonly_df %>%
  distinct(year, locality, scientificName, periodNumber) %>%
  group_by(locality, scientificName) %>%
  summarize(kick_periods_across = n(), .groups = "drop")

malaise_first <- malaise_occurrences_df %>%
  transmute(year, locality, scientificName, periodNumber) %>%
  distinct(year, locality, scientificName, periodNumber) %>%
  group_by(year, locality, scientificName) %>%
  summarize(first_emerged_period = min(periodNumber), .groups = "drop")

first_emergence_df <- full_join(
  kick_periods_year,
  kick_periods_across,
  by = c("locality", "scientificName")
) %>%
  full_join(
    malaise_first,
    by = c("year", "locality", "scientificName")
  ) %>%
  mutate(
    first_emerged = case_when(
      !is.na(first_emerged_period) ~ paste0("Period", first_emerged_period),
      is.na(first_emerged_period) & kick_periods_year >= 2 ~ "Later",
      is.na(first_emerged_period) & kick_periods_across >= 3 ~ "Later",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Period1", "Period2", "Period3", "Period4", "Later")),
    Status = lake_status(locality) %>% factor(levels = c("Control", "Regulated"))
  ) %>%
  filter(!is.na(first_emerged))


```

```{r Overview table of first emergence}

# Make a table with the first emergence of Trichoptera species in Malaise traps
firstemergence_table_wide <- first_emergence_df %>% 
  group_by(year, Status, locality) %>%
  summarise(first_emergence = list(table(first_emerged))) %>%
  unnest_wider(first_emergence)


firstemergence_table_long <- firstemergence_table_wide %>% pivot_longer(
  cols = c("Period1", "Period2", "Period3", "Period4", "Later"), 
  names_to = "first_emergence", 
  values_to = "count"
)

firstemergence_table_wide
firstemergence_table_long

```

```{r A2: CLMMs for first emergence}

# Do a cumulative logit model of first emergence


A2_mod_0_null <- ordinal::clmm(first_emerged ~ 1 + (1|scientificName) + (1|locality),
                                           data = first_emergence_df, 
                                           link = "logit")

A2_mod_1_year <- ordinal::clmm(first_emerged ~ year + (1|scientificName) + (1|locality),
                                           data = first_emergence_df, 
                                           link = "logit")
A2_mod_2_status <- ordinal::clmm(first_emerged ~ Status + (1|scientificName) + (1|locality),
                                           data = first_emergence_df, 
                                           link = "logit")
A2_mod_3_year_status <- ordinal::clmm(first_emerged ~ year + Status + (1|scientificName) + (1|locality),
                                           data = first_emergence_df, 
                                           link = "logit")
A2_mod_4_interaction <- ordinal::clmm(first_emerged ~ year * Status + (1|scientificName) + (1|locality),
                                           data = first_emergence_df, 
                                           link = "logit")



A2_candidate_models <- list(
    A2_mod_0_null,
    A2_mod_1_year,
    A2_mod_2_status,
    A2_mod_3_year_status,
    A2_mod_4_interaction
  )
MuMIn::model.sel(A2_candidate_models)

A2_favourite_model <- MuMIn::get.models(MuMIn::model.sel(A2_candidate_models), subset = 1)[[1]]
A2_favourite_model

summary(A2_favourite_model)
confint(A2_favourite_model)
```

```{r A2: All model summaries}

lapply(A2_candidate_models, summary)

```

```{r A2: Calculating predictions}

# A2_favourite_model has two predictors: Status and Year

# Create a new data frame for prediction (e.g., all combinations of Status and year)
A2_firstemg_predictions <- expand.grid(
  j_first_emergence = levels(first_emergence_df$first_emerged),
  x1_Status = levels(first_emergence_df$Status),
  x2_year = levels(first_emergence_df$year)
)

# Get values for thresholds (alpha) and predictors (beta)
A2_firstemg_predictions <- A2_firstemg_predictions %>% mutate(
  alpha_j = case_when(
    j_first_emergence == "Period2" ~ A2_favourite_model$alpha[["Period2|Period3"]],
    j_first_emergence == "Period3" ~ A2_favourite_model$alpha[["Period3|Period4"]],
    j_first_emergence == "Period4" ~ A2_favourite_model$alpha[["Period4|Later"]]
    ),
  beta_x1 = case_when(
    x1_Status == "Control" ~ 0,
    x1_Status == "Regulated" ~ 1*A2_favourite_model$beta[["StatusRegulated"]]
    ),
  beta_x2 = case_when(
    x2_year == 2023 ~ 0,
    x2_year == 2024 ~ 1*A2_favourite_model$beta[["year2024"]]
    )
  )

# Calculate logit(P(Y<=j))
A2_firstemg_predictions <- A2_firstemg_predictions %>% mutate(logit_P_Y_leq_j = case_when(
  j_first_emergence == "Period1" ~ NA,
  .default = alpha_j - beta_x1 - beta_x2))

# Calculate P(Y<=j)
A2_firstemg_predictions <- A2_firstemg_predictions %>% mutate(exp_logit = exp(logit_P_Y_leq_j))
A2_firstemg_predictions <- A2_firstemg_predictions %>% mutate(P_Y_leq_j = case_when(
  j_first_emergence == "Period1" ~ 0.0,
  j_first_emergence == "Later" ~ 1.0, 
  .default = exp_logit/(1+exp_logit)))

# Calculate P(Y=j)
A2_firstemg_predictions <- A2_firstemg_predictions %>% mutate(P_Y_eq_j = case_when(
  j_first_emergence == "Period1" ~ 0,
  .default = P_Y_leq_j-lag(P_Y_leq_j)))

```

```{r A2: Prediction uncertainty via simulation}
# Simulate coefficient uncertainty for the fitted CLMM and derive prediction intervals.
# This captures uncertainty in thresholds and fixed effects for the marginal predictions.


# For ordinal CLMM, extract both thresholds (alpha) and fixed effects (beta)
coefs <- c(A2_favourite_model$alpha, A2_favourite_model$beta)

vc    <- vcov(A2_favourite_model)

# Ensure vcov is subsetted to match coefs
vc_matched <- vc[names(coefs), names(coefs)]

set.seed(2026)
n_sims <- 2000
sim_coef_mat <- MASS::mvrnorm(n = n_sims, mu = coefs, Sigma = vc_matched)

simulated_A2_predictions <- lapply(seq_len(n_sims), function(sim_i) {
  theta <- sim_coef_mat[sim_i, ]

  A2_firstemg_predictions %>%
    mutate(
      alpha_j = case_when(
        j_first_emergence == "Period2" ~ theta[["Period2|Period3"]],
        j_first_emergence == "Period3" ~ theta[["Period3|Period4"]],
        j_first_emergence == "Period4" ~ theta[["Period4|Later"]]
      ),
      beta_x1 = case_when(
        x1_Status == "Control" ~ 0,
        x1_Status == "Regulated" ~ theta[["StatusRegulated"]]
      ),
      beta_x2 = case_when(
        x2_year == 2023 ~ 0,
        x2_year == 2024 ~ theta[["year2024"]]
      ),
      logit_P_Y_leq_j = alpha_j - beta_x1 - beta_x2,
      P_Y_leq_j = case_when(
        j_first_emergence == "Period1" ~ 0,
        j_first_emergence == "Period2" ~ plogis(logit_P_Y_leq_j),
        j_first_emergence == "Period3" ~ plogis(logit_P_Y_leq_j),
        j_first_emergence == "Period4" ~ plogis(logit_P_Y_leq_j),
        j_first_emergence == "Later" ~ 1
      )
    ) %>%
    group_by(x1_Status, x2_year) %>%
    arrange(j_first_emergence, .by_group = TRUE) %>%
    mutate(P_Y_eq_j = P_Y_leq_j - lag(P_Y_leq_j, default = 0)) %>%
    ungroup() %>%
    mutate(sim = sim_i)
})

simulated_A2_predictions <- dplyr::bind_rows(simulated_A2_predictions)

A2_prediction_intervals <- simulated_A2_predictions %>%
  group_by(j_first_emergence, x1_Status, x2_year) %>%
  summarise(
    P_Y_eq_j_mean = mean(P_Y_eq_j),
    P_Y_eq_j_se   = sd(P_Y_eq_j),
    P_Y_eq_j_lo95   = quantile(P_Y_eq_j, 0.025),
    P_Y_eq_j_hi95   = quantile(P_Y_eq_j, 0.975),
    P_Y_eq_j_lo68   = quantile(P_Y_eq_j, 0.5-0.341),
    P_Y_eq_j_hi68   = quantile(P_Y_eq_j, 0.5+0.341),
    P_Y_leq_j_mean = mean(P_Y_leq_j),
    P_Y_leq_j_se   = sd(P_Y_leq_j),
    P_Y_leq_j_lo95   = quantile(P_Y_leq_j, 0.025),
    P_Y_leq_j_hi95   = quantile(P_Y_leq_j, 0.975),
    P_Y_leq_j_lo68   = quantile(P_Y_leq_j, 0.5-0.341),
    P_Y_leq_j_hi68   = quantile(P_Y_leq_j, 0.5+0.341),
    .groups = "drop"
  )

A2_firstemg_predictions <- A2_firstemg_predictions %>%
  left_join(A2_prediction_intervals,
            by = c("j_first_emergence", "x1_Status", "x2_year"))

```

```{r A2: Visualize predictions with uncertainty}
# Line-and-dot plot with ribbon and error bars for P(Y=j) and cumulative P(Y<=j).

p_eq_A2 <- ggplot(A2_firstemg_predictions, aes(x = j_first_emergence, y = P_Y_eq_j_mean, group = x1_Status, color = x1_Status)) +
  geom_errorbar(aes(ymin = P_Y_eq_j_lo95, ymax = P_Y_eq_j_hi95), width = 0.2, position = position_dodge(width = 0.3)) +
  geom_line(position = position_dodge(width = 0.3), linewidth = 1) +
  geom_point(position = position_dodge(width = 0.3), size = 2) +
  facet_wrap(~x2_year, nrow = 2) +
  labs(
       x = "Sampling Period (j)",
       y = "P(First emergence = j)") +
  scale_color_manual(name = "Status", values = status_colours) +
  scale_fill_manual(name = "Status", values = status_colours) +
  scale_x_discrete(labels = c("Period1" = "1", "Period2" = "2", "Period3" = "3", "Period4" = "4", "Later" = "Later")) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank())
  

p_leq_A2 <- ggplot(A2_firstemg_predictions, aes(x = j_first_emergence, y = P_Y_leq_j_mean, group = x1_Status, color = x1_Status)) +
  geom_errorbar(aes(ymin = P_Y_leq_j_lo95, ymax = P_Y_leq_j_hi95), width = 0.2, position = position_dodge(width = 0.3)) +
  geom_line(position = position_dodge(width = 0.3), linewidth = 1) +
  geom_point(position = position_dodge(width = 0.3), size = 2) +
  facet_wrap(~x2_year, nrow = 2) +
  labs(
       x = "Sampling Period (j)",
       y = "P(First emergence \u2264 j)") +
  scale_color_manual(name = "Status", values = status_colours) +
  scale_fill_manual(name = "Status", values = status_colours) +
  scale_x_discrete(labels = c("Period1" = "1", "Period2" = "2", "Period3" = "3", "Period4" = "4", "Later" = "Later")) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank())

p_eq_A2
p_leq_A2

for (figformat in preferred_formats){
  # NOTE! For some reason "less or equal to" symbol is displayed as = in pdf (svg and png OK)
  ggsave(paste("figures/results/firstemg_peq.", figformat, sep = ""), plot = p_eq_A2, height = 10, width = col_width_dbl, units = "cm")
  ggsave(paste("figures/results/firstemg_pleq.", figformat, sep = ""), , plot = p_leq_A2, height = 10, width = col_width_dbl, units = "cm")
}

  
```

```{r A2: Closer look at species difference between regulated and control}

# Subset for species that are observed in both regulated and control lakes
species_both <- first_emergence_df %>% group_by(scientificName) %>% filter(n_distinct(Status) == 2) %>% pull(scientificName) %>% unique()
first_emergence_df_bothonly <- first_emergence_df %>% filter(scientificName %in% species_both)

species_regonly <- first_emergence_df %>% group_by(scientificName) %>% filter(n_distinct(Status) == 1, Status == "Regulated") %>% pull(scientificName) %>% unique()
first_emergence_df_regonly <- first_emergence_df %>% filter(scientificName %in% species_regonly)


species_ctrlonly <- first_emergence_df %>% group_by(scientificName) %>% filter(n_distinct(Status) == 1, Status == "Control") %>% pull(scientificName) %>% unique()
first_emergence_df_ctrlonly <- first_emergence_df %>% filter(scientificName %in% species_ctrlonly)



# Function classifying each species by which lake types it was recorded in
species_statusgroup <- function(speciesname){
  case_when(
    speciesname %in% species_both ~ "Both",
    speciesname %in% species_regonly ~ "Regulated",
    speciesname %in% species_ctrlonly ~ "Control",
    .default = c("Unknown group")
    )
}


```

```{r Plotting first emergence for each species}

first_emergence_df$species_statusgroup <- species_statusgroup(first_emergence_df$scientificName)

first_emergence_df$species_name <- str_extract(first_emergence_df$scientificName, "^\\w* \\w*")
first_emergence_df$taxa_description <- str_replace(first_emergence_df$scientificName, "^\\w* \\w* ", "")
first_emergence_df$markdown_name <- paste0("<i>", first_emergence_df$species_name,"</i>  ", first_emergence_df$taxa_description)


p_all <- ggplot(first_emergence_df, aes(x = first_emerged, y=markdown_name, fill = Status, shape = Status, alpha = 0.4)) +
  facet_grid(rows = vars(species_statusgroup), cols = vars(year), scales = "free_y", space = "free_y") +
  geom_count() +
  scale_fill_manual(values = status_colours) +
  scale_shape_manual(values = c(21,22)) +
  scale_size_area(breaks = c(1, 2, 3), max_size = 6, transform = "exp", name = "Number of lakes") +
  guides(alpha = "none") +
  xlab("First emergence") +
  scale_y_discrete(limits=rev) +
  theme_bw() +
  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank(), legend.position = "bottom", legend.box="vertical",
        axis.text.y = ggtext::element_markdown())
show(p_all)

for (figformat in preferred_formats){
  ggsave(paste("figures/results/specieslevel_firstemg_all.", figformat, sep = ""), plot = p_all, height = 19, width = col_width_sng, units = "cm")
}


```



# ANALYSIS 3: ABUNDANCE AND SPECIES RICHNESS -------------

The start of a phenological event, such as emergence, can be affected by the number of individuals, and since we are also looking across many species, the number of species as well. Because we only have data from the start of the flight period, we use larval data (kick samples) to calculate Trichoptera abundance and richness.

This data is on sample level, and only includes individuals identified to species level.

```{r Get all kicksample event IDs}
# To get all the kicksamples that have been processed, regardless of whether Trichoptera are present or not, we load the data from the first sorting/identification round.

# Import all occurrence files that are in the folder "occurrences"
kicksample_sorting_df <- list.files(path = here::here("data","source_data","raw","kicksample_occurrences"), # create a list of all excel files in folder
                             pattern = "*xlsx",
                             full.names = TRUE) %>%
  lapply(read_excel, sheet = "grovsortering") %>% # read all the files in the list
  bind_rows() # Merge into single dataframe

# Remove empty rows (where id is NA), result of adding empty rows in excel
kicksample_sorting_df <- kicksample_sorting_df[!is.na(kicksample_sorting_df$internal_parentEventID),] 

# Remove unnecessary columns
kicksample_sorting_df <- kicksample_sorting_df %>% 
  dplyr::select(internal_eventID,scientificName,individualCount) # remove "internal_parentEventID" and "order" columns (should not have been in excel)

# Change "round" to "sample" 
kicksample_sorting_df$internal_eventID <- str_replace_all(kicksample_sorting_df$internal_eventID, "round","sample")

#### Add info ------------

# Add locality/event info to occurrence file
kicksample_sorting_df <- merge(x = kicksample_sorting_df, y = events_df,
                        by = "internal_eventID", all.x = TRUE) 
  # ideally, here we will only have one key column, internal_eventID


# Add taxon info to occurrence file
kicksample_sorting_df <- merge(x = kicksample_sorting_df, y = taxonlist_df,
                        by = "scientificName", all.x = TRUE)

#### Tidy data --------------

# Remove incomplete autumn sampling round
kicksample_sorting_df <- kicksample_sorting_df %>%
  filter(month < 8) 


# Ø -> O 
kicksample_sorting_df <- kicksample_sorting_df %>%
  mutate(internal_parentEventID = iconv(internal_parentEventID,"UTF-8", "ASCII//TRANSLIT", sub = "o"),
         internal_eventID = iconv(internal_eventID,"UTF-8", "ASCII//TRANSLIT", sub = "o"),
         locality = iconv(locality,"UTF-8", "ASCII//TRANSLIT", sub = "o"))



# Create regulation status column 
kicksample_sorting_df <- kicksample_sorting_df %>%
  mutate(Status = lake_status(locality))


# Make locality a factor and set level order 
kicksample_sorting_df$locality <- factor(kicksample_sorting_df$locality, 
                                  levels = lake_names)

# Make year a factor and set level order
kicksample_sorting_df$year <- factor(kicksample_sorting_df$year, 
                              levels = c(2023, 2024))



# Remove additional replicates

# For lake Stor-Drakstsjøen, two additional replicates were semi-processed (EPT was counted, but not identified to species level) for the date 24.05.2023. This was done early in the sorting process, when the time available for labwork was being assessed. Further, having multiple replicates for the same time point allowed us to compare how similar replicates were in terms of species composition and abundance.
# 

kicksample_sorting_df <- kicksample_sorting_df %>%
  dplyr::filter(!internal_eventID %in% c("Stor-Drakstsjoen_24_05_2023_sample_4",
                                       "Stor-Drakstsjoen_24_05_2023_sample_5"))

```

```{r Dataset for abundance and richness analysis}

# To get all 

kicksample_overview_df <- kicksample_sorting_df %>% transmute(internal_eventID, year, Status, locality)
kicksample_overview_df <- kicksample_overview_df[!duplicated(kicksample_overview_df), ] # Should be length 60

kicksample_overview_df <- left_join(kicksample_overview_df, 
                                    kicksample_trich_occurrences_speciesonly_df %>% 
                                      group_by(internal_eventID) %>% 
                                      summarise(sum_individualCount = sum(individualCount), sum_Species = length(unique(speciesKey))),
                                    by = "internal_eventID")

kicksample_overview_df$sum_individualCount <- replace_na(kicksample_overview_df$sum_individualCount, 0)
kicksample_overview_df$sum_Species <- replace_na(kicksample_overview_df$sum_Species, 0)

```


## Quick stats for larvae (kick samples)

```{r Kick samples: Quick stats}
print("---- Stats across all kick samples ----")

paste("Number of individuals (ID): ", sum(kicksample_trich_occurrences_speciesonly_df$individualCount), 
      "(doublecheck: ", sum(kicksample_overview_df$sum_individualCount),")") %>% print()
paste("Number of individuals (ID): ", sum(kicksample_trich_occurrences_speciesonly_df$individualCount)) %>% print()
paste("Number of species: ", length(unique(filter(kicksample_trich_occurrences_speciesonly_df)$scientificName))) %>% print()

print("---- Stats for reg kick samples ----")
paste("Number of individuals (ID): ", sum(filter(kicksample_trich_occurrences_speciesonly_df, Status == "Regulated")$individualCount)) %>% print()
paste("Number of species: ", length(unique(filter(kicksample_trich_occurrences_speciesonly_df, Status == "Regulated")$scientificName))) %>% print()

print("---- Stats for ctrl kick samples ----")
paste("Number of individuals (ID): ", sum(filter(kicksample_trich_occurrences_speciesonly_df, Status == "Control")$individualCount)) %>% print()
paste("Number of species: ", length(unique(filter(kicksample_trich_occurrences_speciesonly_df, Status == "Control")$scientificName))) %>% print()


```


## Abundance

```{r Overview abundance}
# Investigate dataset structure ---------------------------------------------

## Basic stats -------------------------------------------------------------

mean(kicksample_overview_df$sum_individualCount) # 8.016667
var(kicksample_overview_df$sum_individualCount) # 98.66073
sd(kicksample_overview_df$sum_individualCount) # 9.932811
min(kicksample_overview_df$sum_individualCount) # 0
max(kicksample_overview_df$sum_individualCount) # 39


stderror <- function(x) sd(x)/sqrt(length(x))
stderror(kicksample_overview_df$sum_individualCount) # 1.28232

mean(filter(kicksample_overview_df, Status == "Regulated")$sum_individualCount) # 4.3
mean(filter(kicksample_overview_df, Status == "Control")$sum_individualCount) # 11.73

sum(filter(kicksample_overview_df, Status == "Regulated")$sum_individualCount) # 129
sum(filter(kicksample_overview_df, Status == "Control")$sum_individualCount) # 352


## Figures ---------------------------------------------

# Dot plot
p_kickabun_dot <- ggplot(data = kicksample_overview_df, aes(x = Status, y = sum_individualCount, fill = Status)) + 
  geom_dotplot(binaxis = "y", stackdir = "center", color = NA) + 
  ylab("Abundance (ind/sample)") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_fill_manual(values = status_colours) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  guides(fill = "none")

# Box plot
p_kickabun_box <- ggplot(data = kicksample_overview_df, aes(x = Status, y = sum_individualCount, fill = Status, color = Status)) + 
  geom_boxplot() + 
  facet_wrap(~year) +
  ylab("Abundance (ind/sample)") +
  xlab("") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5), expand = expansion(mult = c(0, 0.05))) +
  scale_color_manual(values = status_colours) +
  scale_fill_manual(values = 
                      c("Control" = alpha(status_colours[["Control"]], alpha = 0.5), 
                        "Regulated" = alpha(status_colours[["Regulated"]], alpha = 0.5))) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  guides(fill = "none", color = "none")


for (figformat in preferred_formats){
  ggsave(paste("figures/results/kickAbun_dotplot.", figformat, sep = ""), plot = p_kickabun_dot, height = 7.5, width = col_width_dbl, units = "cm")
  ggsave(paste("figures/results/kickAbun_boxplot.", figformat, sep = ""), plot = p_kickabun_box, height = 7.5, width = col_width_dbl, units = "cm")
}


```

```{r Abundance: GLMMs (negative binomial)}

# Fit candidate models

# Model 1 full 
kicksample_ab_nb_1 <- glmer.nb(formula = sum_individualCount ~ Status*year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)
# NB! Model 1 gives "not converging" warning

# Model 2 no interaction
kicksample_ab_nb_2 <- glmer.nb(formula = sum_individualCount ~ Status + year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 3 only Status
kicksample_ab_nb_3 <- glmer.nb(formula = sum_individualCount ~ Status + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 4 only year
kicksample_ab_nb_4 <- glmer.nb(formula = sum_individualCount ~ year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 5 only intercept
kicksample_ab_nb_5 <- glmer.nb(formula = sum_individualCount ~ 1 + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# List all candidate models
ab_candidate_models <- list(kicksample_ab_nb_1,kicksample_ab_nb_2,kicksample_ab_nb_3,kicksample_ab_nb_4,kicksample_ab_nb_5)
MuMIn::model.sel(ab_candidate_models)

lapply(ab_candidate_models, summary)

# Favourite model
ab_favourite_model <- MuMIn::get.models(MuMIn::model.sel(ab_candidate_models), subset = 1)[[1]]
ab_favourite_model
coef(ab_favourite_model)
summary(ab_favourite_model)

```


## Richness

```{r Overview richness}
# Investigate dataset structure ---------------------------------------------

## Basic stats -------------------------------------------------------------

mean(kicksample_overview_df$sum_Species) # 2.45
var(kicksample_overview_df$sum_Species) # 4.895763
sd(kicksample_overview_df$sum_Species) # 2.212637
min(kicksample_overview_df$sum_Species) # 0
max(kicksample_overview_df$sum_Species) # 8

stderror(kicksample_overview_df$sum_Species) # 0.2856502

mean(filter(kicksample_overview_df, Status == "Regulated")$sum_Species) # 1.733333
mean(filter(kicksample_overview_df, Status == "Control")$sum_Species) # 3.166667

sum(filter(kicksample_overview_df, Status == "Regulated")$sum_Species) # 52
sum(filter(kicksample_overview_df, Status == "Control")$sum_Species) # 95


## Figures ---------------------------------------------

# Dot plot
p_kickrich_dot <- ggplot(data = kicksample_overview_df, aes(x = Status, y = sum_Species, fill = Status)) + 
  geom_dotplot(binaxis = "y", stackdir = "center", color = NA) + 
  ylab("Species richness (species/sample)") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_fill_manual(values = status_colours) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  guides(fill = "none")

# Box plot
p_kickrich_box <- ggplot(data = kicksample_overview_df, aes(x = Status, y = sum_Species, fill = Status, color = Status)) + 
  geom_boxplot() + 
  facet_wrap(~year) +
  ylab("Species richness (species/sample)") +
  xlab("") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5), expand = expansion(mult = c(0, 0.05))) +
  scale_color_manual(values = status_colours) +
  scale_fill_manual(values = 
                      c("Control" = alpha(status_colours[["Control"]], alpha = 0.5), 
                        "Regulated" = alpha(status_colours[["Regulated"]], alpha = 0.5))) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  guides(fill = "none", color = "none")



for (figformat in preferred_formats){
    ggsave(paste("figures/results/kickRich_dotplot.", figformat, sep = ""), plot = p_kickrich_dot, height = 7.5, width = col_width_dbl, units = "cm")
  ggsave(paste("figures/results/kickRich_boxplot.", figformat, sep = ""), plot = p_kickrich_box, height = 7.5, width = col_width_dbl, units = "cm")
}


```

```{r Richness: GLMM (Negative binominal distribution)}
# Fit candidate models 

# All models give warning:
# in theta.ml(Y, mu, weights = object@resp$weights, limit = limit,  :iteration limit reached

# Model 1 full 
kicksample_rich_nb_1 <- glmer.nb(formula = sum_Species ~ Status*year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 2 no interaction
kicksample_rich_nb_2 <- glmer.nb(formula = sum_Species ~ Status + year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 3 only Status
kicksample_rich_nb_3 <- glmer.nb(formula = sum_Species ~ Status + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 4 only year
kicksample_rich_nb_4 <- glmer.nb(formula = sum_Species ~ year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# Model 5 only intercept
kicksample_rich_nb_5 <- glmer.nb(formula = sum_Species ~ 1 + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude)

# List all candidate models
rc_candidate_models <- list(kicksample_rich_nb_1,kicksample_rich_nb_2,kicksample_rich_nb_3,kicksample_rich_nb_4,kicksample_rich_nb_5)
MuMIn::model.sel(rc_candidate_models)

```

```{r Richness: GLMM (Poisson distribution)}

# Fit candidate models 


# Model 1 full 
kicksample_rich_po_1 <- glmer(formula = sum_Species ~ Status*year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude, 
                     family = "poisson")

# Model 2 no interaction
kicksample_rich_po_2 <- glmer(formula = sum_Species ~ Status + year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude, family = "poisson")

# Model 3 only Status
kicksample_rich_po_3 <- glmer(formula = sum_Species ~ Status + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude, family = "poisson")

# Model 4 only year
kicksample_rich_po_4 <- glmer(formula = sum_Species ~ year + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude, family = "poisson")

# Model 5 only intercept
kicksample_rich_po_5 <- glmer(formula = sum_Species ~ 1 + (1|locality), 
                     data = kicksample_overview_df,
                     na.action = na.exclude, family = "poisson")

# List all candidate models
rc_candidate_models <- list(kicksample_rich_po_1,kicksample_rich_po_2,kicksample_rich_po_3,kicksample_rich_po_4,kicksample_rich_po_5)
MuMIn::model.sel(rc_candidate_models)

lapply(rc_candidate_models, summary)

# Favourite model
rc_favourite_model <- MuMIn::get.models(MuMIn::model.sel(rc_candidate_models), subset = 1)[[1]]
rc_favourite_model
coef(kicksample_rich_po_1)
summary(kicksample_rich_po_1)

```



# ADDITIONAL CHECKS ------------

```{r A1: Malaise count table (all Trich individuals)}

# Load files
malaise_sorting_df <- list.files(path = "data/source_data/raw/malaise_data_occurences/", # go to folder
                         pattern = "*.xlsx", # look for Excel-files
                         full.names = TRUE) %>% # include path in filename
  lapply(read_excel, sheet = "grovsortering") %>% # read all the files in the list 
  bind_rows() %>% # Merge into single dataframe
  filter(complete.cases(.)) # remove incomplete cases, i.e. rows with NA

# Remove unnecessary/duplicate columns
malaise_sorting_df <- malaise_sorting_df %>% 
  dplyr::select(internal_eventID,scientificName,individualCount) # remove "internal_parentEventID" and "order" columns (should not have been in excel)

# Change "round" to "sample" 
malaise_sorting_df$internal_eventID <- str_replace_all(malaise_sorting_df$internal_eventID, "round","sample")

# Add locality/event info to occurrence file
malaise_sorting_df <- merge(x = malaise_sorting_df, y = events_df,
                    by = c("internal_eventID"), all.x = TRUE) 

# Update to correct protocol
malaise_sorting_df$samplingProtocol <- "Malaise trap"
malaise_sorting_df$samplingEffort <- "Single Malaise trap, operated for approximately two weeks"

# Add taxon info to occurrence file
malaise_sorting_df <- merge(x = malaise_sorting_df, y = taxonlist_df,
                    by = "scientificName", all.x = TRUE)

# Change ø to o
malaise_sorting_df <- malaise_sorting_df %>%
  mutate(locality = iconv(locality,"UTF-8", "ASCII//TRANSLIT", sub = "o"),
         internal_eventID = iconv(internal_eventID,"UTF-8", "ASCII//TRANSLIT", sub = "o"))

# Create regulation status column
malaise_sorting_df <- malaise_sorting_df %>%
  mutate(Status = lake_status(locality))


# Make locality a factor and set level order
malaise_sorting_df$locality <- factor(malaise_sorting_df$locality, 
                                              levels = lake_names)

# Make year a factor and set level order
malaise_sorting_df$year <- factor(malaise_sorting_df$year, levels = c(2023, 2024))

# Subset for only Trichoptera
malaise_sorting_T_df <- malaise_sorting_df %>% filter(scientificName == "Trichoptera")



# Table 
A1_adultcount2_table_long <- malaise_sorting_T_df %>% 
  group_by(year, Status, locality, periodNumber) %>%
  summarise(sum_n = sum(individualCount)) %>%
  ungroup() %>%
  complete(year, locality, periodNumber = c(1,2,3,4)) %>%
  transmute(year, Status, locality, periodNumber, sum_n) %>% 
  mutate(Status = lake_status(locality))


# Pivot to wide format (each period number as a column)
A1_adultcount2_table_wide <- A1_adultcount2_table_long %>%
  pivot_wider(names_from = periodNumber, values_from = sum_n, names_prefix = "Period")

A1_adultcount2_table_wide

```

```{r Malaise: Compare N ID'd to N all}
print("---- Stats across all Malaise ----")
paste("Number of individuals (ID): ", sum(malaise_occurrences_df$individualCount)) %>% print()
paste("Number of individuals (all): ", sum(malaise_sorting_T_df$individualCount)) %>% print()

print("---- Stats for reg Malaise ----")
paste("Number of individuals (ID): ", sum(filter(malaise_occurrences_df, Status == "Regulated")$individualCount)) %>% print()
paste("Number of individuals (all): ", sum(filter(malaise_sorting_T_df, Status == "Regulated")$individualCount)) %>% print()

print("---- Stats for ctrl Malaise ----")
paste("Number of individuals (ID): ", sum(filter(malaise_occurrences_df, Status == "Control")$individualCount)) %>% print()
paste("Number of individuals (all): ", sum(filter(malaise_sorting_T_df, Status == "Control")$individualCount)) %>% print()

```

```{r Kick samples: Compare N ID'd to N all}
print("---- Stats across all kick samples ----")

paste("Number of individuals (ID): ", sum(filter(kicksample_trich_occurrences_speciesonly_df, taxonRank == "SPECIES")$individualCount)) %>% print()
paste("Number of individuals (ID): ", sum(kicksample_overview_df$sum_individualCount)) %>% print()
paste("Number of individuals (all): ", sum(filter(kicksample_sorting_df, order == "Trichoptera")$individualCount)) %>% print()

print("---- Stats for reg kick samples ----")
paste("Number of individuals (ID): ", sum(filter(kicksample_trich_occurrences_speciesonly_df, taxonRank == "SPECIES", Status == "Regulated")$individualCount)) %>% print()
paste("Number of individuals (ID): ", sum(filter(kicksample_overview_df, Status == "Regulated")$sum_individualCount)) %>% print()
paste("Number of individuals (all): ", sum(filter(kicksample_sorting_df, order == "Trichoptera", Status == "Regulated")$individualCount)) %>% print()

print("---- Stats for ctrl kick samples ----")
paste("Number of individuals (ID): ", sum(filter(kicksample_trich_occurrences_speciesonly_df, taxonRank == "SPECIES", Status == "Control")$individualCount)) %>% print()
paste("Number of individuals (ID): ", sum(filter(kicksample_overview_df, Status == "Control")$sum_individualCount)) %>% print()
paste("Number of individuals (all): ", sum(filter(kicksample_sorting_df, order == "Trichoptera", Status == "Control")$individualCount)) %>% print()

```
