---
title: "temperature"
output: html_document
---

##############################################

# 2.3 Temperature data

##############################################

# Import temperature data from HOBO loggers

# Setup ------------------------------------------------------------------------

```{r}
colnames_temperature <- c("datapoint_nr","date_time","temperature_C","logger_id","logger_nr")

source("figure_formats_and_colour_palette.R")

```

# DATA IMPORT

Description: In the HOBOware pro, settings to data export set to export file type == .xlsx, include line number column, include plot title in header, no quotes or commas in headings..., date format == YMD, date separator == dash, time format == 24-hour. Time zone is GMT +2

The excel files are manually edited: make a new column with the plot title as logger id, and a new column with logger number. Delete first row.

```{r Import excel files}
# 2023 -------------------------------------------------------------------------
t1 <- readxl::read_excel(path = here("data","source_data","temperature","Barsetvatn_2023_bl_21508213.xlsx"))
colnames(t1) <- colnames_temperature
t1 <- t1 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])
  
t2 <- readxl::read_excel(path = here("data","source_data","temperature","Barsetvatn_2023_str_21140198.xlsx"))
colnames(t2) <- colnames_temperature
t2 <- t2 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t3 <- readxl::read_excel(path = here("data","source_data","temperature","Gjoljavatn_2023_bl_21508207.xlsx"))
colnames(t3) <- colnames_temperature
t3 <- t3 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t4 <- readxl::read_excel(path = here("data","source_data","temperature","Gjoljavatn_2023_str_21140205.xlsx"))
colnames(t4) <- colnames_temperature
t4 <- t4 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t5 <- readxl::read_excel(path = here("data","source_data","temperature","Kilvatn_2023_bl_21620476.xlsx"))
colnames(t5) <- colnames_temperature
t5 <- t5 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t6 <- readxl::read_excel(path = here("data","source_data","temperature","Kilvatn_2023_str_21140206.xlsx"))
colnames(t6) <- colnames_temperature
t6 <- t6 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t7 <- readxl::read_excel(path = here("data","source_data","temperature","Roksetvatn_2023_str_21508212.xlsx"))
colnames(t7) <- colnames_temperature
t7 <- t7 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t8 <- readxl::read_excel(path = here("data","source_data","temperature","Stordrakstsj_2023_str_21102653.xlsx"))
colnames(t8) <- colnames_temperature
t8 <- t8 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t9 <- readxl::read_excel(path = here("data","source_data","temperature","Storvatn_2023_bl_21508214.xlsx"))
colnames(t9) <- colnames_temperature
t9 <- t9 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t10 <- readxl::read_excel(path = here("data","source_data","temperature","Storvatn_2023_str_21140212.xlsx"))
colnames(t10) <- colnames_temperature
t10 <- t10 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])


# 2024 -------------------------------------------------------------------------

t11 <- readxl::read_excel(path = here("data","source_data","temperature","Barsetvatnet_2024_bl_21774703.xlsx"))
colnames(t11) <- colnames_temperature
t11 <- t11 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t12 <- readxl::read_excel(path = here("data","source_data","temperature","Barsetvatnet_2024_str_21774702.xlsx"))
colnames(t12) <- colnames_temperature
t12 <- t12 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t13 <- readxl::read_excel(path = here("data","source_data","temperature","Gjoljavatnet_2024_bl_21774713.xlsx"))
colnames(t13) <- colnames_temperature
t13 <- t13 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t14 <- readxl::read_excel(path = here("data","source_data","temperature","Gjoljavatnet_2024_str_21798976.xlsx"))
colnames(t14) <- colnames_temperature
t14 <- t14 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t15 <- readxl::read_excel(path = here("data","source_data","temperature","Kilvatn_2024_str_21774708.xlsx"))
colnames(t15) <- colnames_temperature
t15 <- t15 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t16 <- readxl::read_excel(path = here("data","source_data","temperature","Roksetvatn_2024_bl_21798966.xlsx"))
colnames(t16) <- colnames_temperature
t16 <- t16 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t17 <- readxl::read_excel(path = here("data","source_data","temperature","Roksetvatn_2024_str_21798975.xlsx"))
colnames(t17) <- colnames_temperature
t17 <- t17 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t18 <- readxl::read_excel(path = here("data","source_data","temperature","Stordrakstsj_2024_bl_21774707.xlsx"))
colnames(t18) <- colnames_temperature
t18 <- t18 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t19 <- readxl::read_excel(path = here("data","source_data","temperature","Stordrakstsj_2024_str_21774696.xlsx"))
colnames(t19) <- colnames_temperature
t19 <- t19 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t20 <- readxl::read_excel(path = here("data","source_data","temperature","Storvatn_2024_bl_21774695.xlsx"))
colnames(t20) <- colnames_temperature
t20 <- t20 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])

t21 <- readxl::read_excel(path = here("data","source_data","temperature","Storvatnet_2024_str_21774704.xlsx"))
colnames(t21) <- colnames_temperature
t21 <- t21 %>%
  mutate(logger_id = logger_id[1],
         logger_nr = logger_nr[1])


# Merge dataframes ----
temperature_data <- rbind(t1,t2,t3,t4,t5,t6,t7,t8,t9,t10,t11,t12,t13,t14,t15,t16,t17,t18,t19,t20,t21)

# Check datetime format
# temperature_data <- temperature_data %>%
#  as.POSIXct(date_time, format = "%Y.%m.%d %H:%M:%S")

# Save file
save(temperature_data, file = here::here("data","source_data","temperature","temperature_data.rda"))

```



# Add logger info --------------------------------------------------------------
```{r Load logger info}
# Import excel sheet with logging period info
logger_info <- readxl::read_excel(path = here("data","source_data","temperature","HOBO_loggers_overview.xlsx"),
                                  sheet = "logger_info") # relevant columns: date_time_start and date_time_stop


# Each logger was placed in either 1) water column with a buoy (ca. 1-1.5m depth), or 2) littoral zone (ca. 0.5-1.0m depth)
logger_info <- logger_info %>% mutate(logger_placement = case_when(
  str_detect(logger_id, "_str") ~ "Littoral zone",
  str_detect(logger_id, "_bl") ~ "Buoy",
  .default = NA
))

```


```{r Keep only valid logging period (v2)}

# Filter the start and stop datetimes: for each logger, keep only datapoints within the date_time_start and date_time_stop interval

# Add start and stop 
temperature_data_v2 <- left_join(
  x = temperature_data, 
  y = dplyr::select(logger_info, c("locality","date_time_start","date_time_stop","logger_id", "logger_placement")),
  by = "logger_id")


# Filter datapoints, go from 16227 to 12918 observations
temperature_data_v2 <- temperature_data_v2 %>%
  subset(date_time >= date_time_start & date_time <= date_time_stop)

# Remove Kilvatnet Buoy 2023 as this was dragged under, and not at the same depth as the others (go from 12918 to 12037 observations)
temperature_data_v2 <- temperature_data_v2 %>% filter(logger_id != "Kilvatn_2023_bl_21620476")


# Set factors
temperature_data_v2$locality <- factor(temperature_data_v2$locality, 
                                       levels = c("Barsetvatnet","Roksetvatnet","Kilvatnet",
                                                  "Gjoljavatnet","Storvatnet","Stor-Drakstsjoen"))
temperature_data_v2$logger_placement <- as.factor(temperature_data_v2$logger_placement)

```


# Add daily mean temp. ---------------------------------------------------------
```{r}
# Add date as separate column
temperature_data_v2 <- temperature_data_v2 %>%
  mutate(eventDate = as.Date(date_time), .after = date_time,
         year = as.numeric(format(date_time, "%Y")),
         month = as.numeric(format(date_time, "%m")),
         day = as.numeric(format(date_time, "%d")),
         month_day = paste(strftime(date_time, format = "%m-%d"),sep="-")) 

# Set factor
temperature_data_v2$year <- as.factor(temperature_data_v2$year)

# Save
save(temperature_data_v2, file = here::here("data","derived_data","temperature","temperature_data_v2.rda"))


# Add regulation status
temperature_data_v2 <- temperature_data_v2 %>%
  mutate(Status = case_when(locality %in% c("Barsetvatnet","Kilvatnet","Roksetvatnet") ~ "Control", 
                            locality %in% c("Gjoljavatnet","Stor-Drakstsjoen","Storvatnet") ~ "Regulated"))
temperature_data_v2$Status <- as.factor(temperature_data_v2$Status)

           
# Average daily temperature
temperature_data_avg <- temperature_data_v2 %>%
  group_by(Status, locality, logger_placement, year, eventDate, logger_id) %>% summarise(
    daily_avg_temp = round(mean(temperature_C, ),digits = 3),
    daily_avg_temp_sd = round(sd(temperature_C),digits = 3))

```


# Plot temperature -------------------------------------------------------------

```{r}

temperature_data_avg$shape_group <- lake_shapegroup(temperature_data_avg$locality)

p_temperature <- ggplot(temperature_data_avg,
                          aes(x = as.Date(paste(2025, strftime(eventDate, "%m-%d"), sep = "-")),
                              y = daily_avg_temp,
                              color = Status,
                              shape = shape_group,
                              alpha = 0.8)) +
  facet_grid(rows = vars(logger_placement), cols = vars(year)) +
  geom_point(size = 1.5) +
  geom_line() +
  scale_color_manual(values = status_colours) +
  scale_shape_manual(name = "Lake", values = lake_shapegroup_symbols) +
  scale_x_date(breaks = month_breaks, labels = month_labels_3L) +
  ylim(0,NA) +
  labs(x = "Time of year", y = "Water temperature (\u00B0C)") +
  guides(alpha = "none") +
  theme_bw() +
  theme(legend.position = "bottom", legend.box="vertical", panel.grid = element_blank())

p_temperature

for (figformat in preferred_formats){
  ggsave(paste("figures/temperature/all_temperatures.", figformat, sep = ""), plot = p_temperature, height = col_width_sng, width = col_width_sng, units = "cm")
}


```

